Convolutional neural networks performing a visual search task show attention-like limits on accuracy when trained to generalize across multiple search stimuli

### 2019-06-06 docs/ccneuro2019/paper.tex
Updates from Overleaf

\section{Introduction}
\subsection{What are the factors that limit visual search?}
What limits our ability to find what we are looking for in the cluttered noisy world we 
see around us? One of the principle tasks that has been used to investigate this question
is a classic form of visual search task, in which subjects view a stimulus on each trial, 
made up of discrete sets of items on a solid-colored background, and report whether a 
target is present among distractors [@wolfeVisualSearch1998]. 
Most studies experimentally manipulate factors such as features of the targets and 
distractors in order to identify the factors limit visual search 
[@ecksteinVisualSearchRetrospective2011; @wolfeFiveFactorsThat2017].
Importantly, it is these factors that *limit* visual search performance that form the 
basis of seminal theories of selective visual attention, although authors differ in the 
extent to which they identify limitations on visual search as synonymous with selective 
attention. To remain neutral, we will use the term "capacity limitations". 
As we review briefly below for context, there are essentially two competing proposed 
sources of capacity limitations [@palmerSignalDetectionEvidence2011]. 
The results supporting both remain unreconciled, in spite of nearly half a century of 
studies based on this task. The core question is: to 
what extent can limitations on visual search be attributed to an attention-like 
computation, e.g. binding features into items, and to what extent can those limitations be 
attributed to other computations, e.g. a decision-making process subject to  
noisy internal representations?
To foreshadow our approach below, we suggest that another way of posing this question is 
to ask: if some algorithm could produce a statistical model capable of learning from data 
to classify visual search stimuli as "target present" or "target absent" with high 
accuracy, would the accuracy still be subject to some ceiling, due simply to the 
constraints of the task? This way of posing the question bears some similarity to a 
well-established framework in vision research known as ideal observer analsyis 
[@geislerIdealObserverAnalysis2003]. However, few ideal observer models have take the 
form of "pixel-in, behavior-out", and the few that have do not generalize to many 
different types of visual search stimuli, so their predictive power is limited 
[@geislerModelsOvertAttention2011].

Before we discuss how to address this gap in models, we briefly review the two competing 
proposed sources of capacity limitations, and describe how they depend in part on the 
way the visual search task is performed. Adopting the terms of other researchers, we will 
call the first family "attention-limited", and the second "noise-limited" 
[@palmerSignalDetectionEvidence2011]. Both attention-limited and noise-limited models 
depend on what are known as *set-size effects* seem when using visual search stimuli, 
which we will define in terms of how the visual search task is carried out. 
Attention-limited models posit two-stage theories of visual processing, a first 
*preattentive* stage which can process single features, such as color or orientation, 
in parallel, and a second *attentive* stage that binds features to objects that processes 
serially, which results in a bottleneck [@treismanFeatureintegrationAttention1980; 
@wolfeGuidedSearchRevised1994]. Evidence for this theory came from visual search 
experiments where subjects were shown stimuli until responding, and their reaction time 
was measured. Plotting reaction time as a function of set size revealed 
lines whose slope were near zero when a single feature distinguished the target from 
distractors (e.g., "red" v. "green"). The slope of reaction time v. set size increased 
sharply, though, when distinguishing the target from distractors required finding a 
conjunction of features (such as "red vertical" v. "red horizontal or green vertical"). 
This sort of set size effect has been replicated hundreds of times, although it remains 
unclear why some search stimuli seem to be processed effectively in parallel while others 
face a bottleneck [@wolfeWhatCanMillion1998; @wolfeFiveFactorsThat2017].
The second family of capacity limitations can be conceived of as completely parallel but 
noisy processing. These models arose in large part in reaction 
to the way the visual search task was carried out when investigating
serial mechanisms. Early studies that measured reaction times left several factors 
uncontrolled, such as target-distractor similarity [@duncanVisualSearchStimulus1989], 
drops in acuity outside the fovea, eye movements, and effects resulting from visual 
crowding [@bergenRapidDiscriminationVisual1983; @ecksteinLowerVisualSearch1998]. 
Hence researchers designed versions of the visual search task that controlled for such 
factors [@palmerPsychophysicsVisualSearch2000]. Crucially, they showed subjects the 
stimulus only briefly, to prevent eye movements, and measured accuracy instead of reaction
time. Computational models of parallel mechanisms succesfully explained results from 
single feature and conjunction search stimuli. These models borrowed ideas from signal 
detection theory that psychophysicists often employ, where sensory representations are
imagined as distributions, and decisions are seen as thresholds on those distributions.
What both attention-limited and noise-limited models have in common is that they are 
both highly abstracted models of the visual system, in which low-level features pass 
through a hierarchy until reaching a final stage consisting of a simple decision rule.

\subsection{CNNs as models of the visual system?}
Similarly, the architecture of convolutional neural networks (CNNs) now routinely used for
computer vision tasks represents a highly abstracted view of the visual system.
Many researchers have drawn parallels between the architecture of CNNs and the 
architecture of the visual system in the brain [@kriegeskorteDeepNeuralNetworks2015], 
in particular in humans and other primates where this system has been most thoroughly 
studied. Like CNNs, the visual system has a hierarchical structure, and is thought to 
function in part by performing transformations at each level of this hierarchy so that 
high-dimensional, low-level features are mapped into low-dimensional abstract 
representations. Several recent studies in the realm of cognitive neuroscience 
have measured representations in more directly tested the similarities between CNNs and 
the visual system, by comparing representations in both [@khaligh-razaviDeepSupervisedNot2014; @yaminsPerformanceoptimizedHierarchicalModels2014; @yaminsUsingGoaldrivenDeep2016]. Representations 
can be easily extracted from CNNs because they are "transparent box" models--i.e. we have 
access to the "activity" of the individual units from which these models are built when 
we "show" the CNNs a stimulus.
A natural question to ask, for someone interested in visual search and models of the visual 
system, is how CNNs perform this task. If CNNs are excellent models of the visual system, 
one might expect them to be subject to the same limitations on visual search. Alternatively, 
if CNNs are simply function approximators, one might expect that they would be able to 
very easily learn to map the exclusive-or function of "target present or target absent", 
especially when targets and distractors are easily discriminated by single feature 
dimensions such as color. While there have been previous studies of neural networks 
performing visual search tasks, we are aware of only one that employed architectures 
which have been used in studies of the visual system referenced above.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\columnwidth]{fig1/fig1.eps}
\end{center}
\caption{A classic form of visual search task.} 
\label{sample-figure}
\end{figure}

>>>>

\section{Introduction}
\subsection{What are the factors that limit visual search?}
What limits our ability to find what we are looking for in the cluttered noisy world we 
see around us? One of the principle tasks that has been used to investigate this question
is a classic form of visual search task (Figure~\ref{fig1}), in which subjects view a stimulus on each trial, 
made up of discrete sets of items on a solid-colored background, and report whether a 
target is present among distractors [@wolfeVisualSearch1998]. 
Most studies experimentally manipulate factors such as features of the targets and 
distractors in order to identify the factors limit visual search 
[@ecksteinVisualSearchRetrospective2011; @wolfeFiveFactorsThat2017].
Importantly, it is these factors that *limit* visual search performance that form the 
basis of seminal theories of selective visual attention, although authors differ in the 
extent to which they identify limitations on visual search as synonymous with selective 
attention. To remain neutral, we will use the term "capacity limitations". 
As we review briefly below for context, there are essentially two competing proposed 
sources of capacity limitations [@palmerSignalDetectionEvidence2011]. 
The results supporting both remain unreconciled, in spite of nearly half a century of 
studies based on this task. The core question is: to 
what extent can limitations on visual search be attributed to an attention-like 
computation, e.g. binding features into items, and to what extent can those limitations be 
attributed to other computations, e.g. a decision-making process subject to  
noisy internal representations?
To foreshadow our approach below, we suggest that another way of posing this question is 
to ask: if some algorithm could produce a statistical model capable of learning from data 
to classify visual search stimuli as "target present" or "target absent" with high 
accuracy, would the accuracy still be subject to some ceiling, due simply to the 
constraints of the task? This way of posing the question bears some similarity to a 
well-established framework in vision research known as ideal observer analsyis 
[@geislerIdealObserverAnalysis2003]. However, few ideal observer models have take the 
form of "pixel-in, behavior-out", and the few that have do not generalize to many 
different types of visual search stimuli, so their predictive power is limited 
[@geislerModelsOvertAttention2011].

Before we discuss how to address this gap in models, we briefly review the two competing 
proposed sources of capacity limitations, and describe how they depend in part on the 
way the visual search task is performed. Adopting the terms of other researchers, we will 
call the first family "attention-limited", and the second "noise-limited" 
[@palmerSignalDetectionEvidence2011]. Both attention-limited and noise-limited models 
depend on what are known as *set-size effects* seem when using visual search stimuli, 
which we will define in terms of how the visual search task is carried out. 
Attention-limited models posit two-stage theories of visual processing, a first 
*preattentive* stage which can process single features, such as color or orientation, 
in parallel, and a second *attentive* stage that binds features to objects that processes 
serially, which results in a bottleneck [@treismanFeatureintegrationAttention1980; 
@wolfeGuidedSearchRevised1994]. Evidence for this theory came from visual search 
experiments where subjects were shown stimuli until responding, and their reaction time 
was measured. Plotting reaction time as a function of set size revealed 
lines whose slope were near zero when a single feature distinguished the target from 
distractors (e.g., "red" v. "green"). The slope of reaction time v. set size increased 
sharply, though, when distinguishing the target from distractors required finding a 
conjunction of features (such as "red vertical" v. "red horizontal or green vertical"). 
This sort of set size effect has been replicated hundreds of times, although it remains 
unclear why some search stimuli seem to be processed effectively in parallel while others 
face a bottleneck [@wolfeWhatCanMillion1998; @wolfeFiveFactorsThat2017].
The second family of capacity limitations can be conceived of as completely parallel but 
noisy processing. These models arose in large part in reaction 
to the way the visual search task was carried out when investigating
serial mechanisms. Early studies that measured reaction times left several factors 
uncontrolled, such as target-distractor similarity [@duncanVisualSearchStimulus1989], 
drops in acuity outside the fovea, eye movements, and effects resulting from visual 
crowding [@bergenRapidDiscriminationVisual1983; @ecksteinLowerVisualSearch1998]. 
Hence researchers designed versions of the visual search task that controlled for such 
factors [@palmerPsychophysicsVisualSearch2000]. Crucially, they showed subjects the 
stimulus only briefly, to prevent eye movements, and measured accuracy instead of reaction
time. Computational models of parallel mechanisms successfully explained results from 
single feature and conjunction search stimuli. These models borrowed ideas from signal 
detection theory that psychophysicists often employ, where sensory representations are
imagined as distributions, and decisions are seen as thresholds on those distributions.
What both attention-limited and noise-limited models have in common is that they are 
both highly abstracted models of the visual system, in which low-level features pass 
through a hierarchy until reaching a final stage consisting of a simple decision rule.

\subsection{CNNs as models of the visual system}
Similarly, the architecture of convolutional neural networks (CNNs) now routinely used for
computer vision tasks represents a highly abstracted view of the visual system.
Many researchers have drawn parallels between the architecture of CNNs and the 
architecture of the visual system in the brain [@kriegeskorteDeepNeuralNetworks2015], 
in particular in humans and other primates where this system has been most thoroughly 
studied. Like CNNs, the visual system has a hierarchical structure, and is thought to 
function in part by performing transformations at each level of this hierarchy so that 
high-dimensional, low-level features are mapped into low-dimensional abstract 
representations. Several recent studies in the realm of cognitive neuroscience 
have measured representations in more directly tested the similarities between CNNs and 
the visual system, by comparing representations in both [@khaligh-razaviDeepSupervisedNot2014; @yaminsPerformanceoptimizedHierarchicalModels2014; @yaminsUsingGoaldrivenDeep2016]. Representations 
can be easily extracted from CNNs because they are "transparent box" models--i.e. we have 
access to the "activity" of the individual units from which these models are built when 
we "show" the CNNs a stimulus.
A natural question to ask, for someone interested in visual search and models of the visual 
system, is how CNNs perform this task. If CNNs are excellent models of the visual system, 
one might expect them to be subject to the same limitations on visual search. Alternatively, 
if CNNs are simply function approximators, one might expect that they would be able to 
very easily learn to map the exclusive-or function of "target present or target absent", 
especially when targets and distractors are easily discriminated by single feature 
dimensions such as color. While there have been previous studies of neural networks 
performing visual search tasks, we are aware of only one that employed architectures 
which have been used in studies of the visual system referenced above.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\columnwidth]{fig1/fig1.eps}
\end{center}
\caption{A classic form of visual search task.} 
\label{fig1}
\end{figure}
